Periodograms for Multi-band Astronomical Time Series

### 2014-12-17 writeup/paper.tex
take a pass at the paper

\title{Multiband Periodograms of Astronomical Sources}

@@

The detection and quantification of periodicity in time-varying signals is an important area of data analysis within modern astronomical surveys.
For evenly-spaced data, the {\it Schuster periodogram}, introduced in 1905, gives a quantitative measure of the periodicity of data as a function of the angular frequency $\omega$. For data $\{d_k\}_{k=1}^N$ measured at equal intervals $t_k = t_0 + k\Delta t$, the Schuster periodogram, which measures the spectral power as a function of the angular frequency, is given by
\begin{equation}
  \eqlabel{Schuster}
  C(\omega) = \frac{1}{N}\left| \sum_{k=1}^N d_k e^{i\omega t_k} \right|^2,
\end{equation}
and can be computed very efficiently using the Fast Fourier Transform.

Because astronomical observing cadences are rarely so uniform, many have looked at extending the ideas behind the periodogram to work with unevenly-sampled data. Most famously, \citep{Lomb76} and \citep{Scargle82} extended earlier work to define the {\it normalized periodogram}:
\begin{equation}
  \eqlabel{LombScargle}
  P_N(\omega) = \frac{1}{2\sigma^2}\left[
    \frac{\left[\sum_k(d_k - \bar{d})\cos\omega(t_k - \tau)\right]^2}
    {\sum_k \cos^2\omega(t_k - \tau)}
    +
    \frac{\left[\sum_k(d_k - \bar{d})\sin\omega(t_k - \tau)\right]^2}
    {\sum_k \sin^2\omega(t_k - \tau)}
\right],
\end{equation}
where $\bar{d}$ is the mean and $\sigma^2$ is the variance of the data $\{d_k\}$, and $\tau$ is the time-offset which makes $P_N(\omega)$ independent of a translation in $t$ \citep[See][for an in-depth discussion]{NumRec}. \citep{Lomb76} showed that this offset has a more important effect: namely, it makes $P_N$ identical to the estimate of harmonic content given a least-squares fit to a single-component sinusoidal model,

@@

This long-recognized connection between spectral power and least squares fitting methods was solidified by \citep{Jaynes87}, who demonstrated that the normalized periodogram of Lomb and Scargle is a sufficient statistic for inferences about a stationary-frequency signal in the presence of Gaussian noise. Building on this result, \citep{Bretthorst88} explored the extension of these methods to more complicated models with multiple frequency terms, nonstationary frequencies, and other more sophisticated models within a Bayesian framework.

@@

In this section we will give a quick quantitative introduction to the least squares fitting formulation of the normalized periodogram of \eq{LombScargle}. We'll denote our $N$ observed data points as
\begin{equation}
  D = \{t_k, y_k, \sigma_k\}_{k=1}^N
\end{equation}
where $t_k$ is the time of observation, $y_k$ is the observed value (typically a magnitude), and $\sigma_k$ describes the Gaussian errors on each value. Without loss of generality we will assume that the data $y_k$ are centered such that the measurements satisfy
\begin{equation}
  \eqlabel{ycentered}
  \frac{\sum_k w_ky_k}{\sum_k w_k} = 0
\end{equation}
where the weights are $w_k = \sigma_k^{-2}$.

@@

The normalized periodogram of \eq{LombScargle} can be derived from the normalized $\chi^2$ of the best-fit single-term stationary sinusoidal model given in \eq{SingleModel}. To make the problem linear, we can re-express the model in terms of the parameter vector $\theta = [A\cos\phi, A\sin\phi]$ so that our model is
\begin{equation}
  \eqlabel{simplemodel}
  y(t|\omega,\theta) = \theta_1\sin(\omega t) + \theta_2\cos(\omega t).
\end{equation}
We can find the maximum likelihood estimate of the parameters $\theta$ by minimizing the $\chi^2$ of the model, which is given by
\begin{equation}
  \chi^2(\omega) = \sum_k \frac{[y_k - y(t_k|\omega,\theta)]^2}{2\sigma_k^2}.
\end{equation}
For the single-term Fourier model, it can be shown \citep[See, e.g.][]{ICVG2014} that
\begin{equation}
  \eqlabel{chi2PN}
  \chi_{min}^2(\omega) = \chi^2_0[1 - P_N(\omega)]
\end{equation}
where $P_N(\omega)$ is the normalized periodogram given in \eq{LombScargle} and $\chi^2_0$ is the reference $\chi^2$ for a constant model, which due to the assumption in \eq{ycentered} is simply $\chi^2_0 = \sum_k (y_k/\sigma_k)^2$

\subsubsection{Matrix Formalism}

These quantities can be expressed more compactly by defining the following matrices:

\begin{equation}
X_\omega = \left[
\begin{array}{cc}
\sin(\omega t_1) & \cos(\omega t_1)\\
\sin(\omega t_2) & \cos(\omega t_2)\\
\vdots & \vdots \\
\sin(\omega t_N) & \cos(\omega t_N)\\
\end{array}
\right];~~
y = \left[
\begin{array}{c}
y_1 \\
y_2\\
\vdots \\
y_N\\
\end{array}
\right];~~
\Sigma = \left[
\begin{array}{cccc}
\sigma_1^2 & 0 &  \cdots & 0\\
0 & \sigma_2^2 &  \cdots & 0\\
\vdots & \vdots &  \ddots & \vdots\\
0 & 0 &  \cdots & \sigma_N^2
\end{array}
\right]
\end{equation}

With these definitions, the model in \eq{simplemodel} is a simple linear product: $y(t|\omega,\theta) = X_\omega\theta$ and the model and reference $\chi^2$ can be written

\begin{eqnarray}
  \chi^2(\omega) &=& (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta)\\
  \chi^2_0 &=& y^T \Sigma^{-1} y
\end{eqnarray}
The normalized periodogram can be computed by finding the minimum of $\chi^2(\omega)$ via standard methods and plugging the result into \eq{chi2PN} to find
\begin{equation}
  \eqlabel{LombScargle2}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
This expression is equivalent to \eq{LombScargle} in the case that $\Sigma \propto I$.

\fig{basic_example} shows an example periodogram for a simulated RR Lyrae light curve based on a template from \citet{Sesar2010}. The particular RR Lyrae star has a period of 0.622 days, and the observations take place on 60 random nights over a 6-month period.  The left panel of the figure shows the input data, and the right panels show the result of a simple Lomb-Scargle periodogram on the data.

@@

In the upper right, we see the normalized periodogram as a function of period. While the power does peak at the true period of 0.622 days, an aliasing effect is apparent. These additional peaks are due to beat frequencies between the true period $P$ and the observing cadence of $\sim 1$ day. That is, for nightly observations, we'd expect excess power to be found at frequencies $P / (1 + nP)$ for a positive or negative integer $n$. This explains the prominent peak at $0.383$. For the current work, we'll ignore this (important) piece of the period-finding problem; for a thorough discussion of aliasing within astronomical time series, see \citet{Roberts87}.

The lower-right panel shows the maximum likelihood interpretation of this periodogram: it is a measure of the normalized $\chi^2$ for a single-term sinusoidal model. Here we visualize the data from the left panel, but folded as a function of phase with the best-fit single-term model over-plot. Here its apparent that the single-term model is biased: RR Lyrae are much more complicated than a simple sinusoid. Nevertheless, the simplistic sinusoidal model does well to zero-in

\subsubsection{Discussion}
We have shown two forms of the classic normalized periodogram: \eq{LombScargle} and \eq{LombScargle2}. Though the two expressions are equivalent, they each have distinct advantages. The expression in \eq{LombScargle} avoids the explicit construction of a matrix, and thus can be computed very efficiently. In particular, through some computational tricks involving the FFT, expressions of the form of \eq{LombScargle} can be evaluated for $N$ frequencies in $\mathcal{O}[\log{N}]$ time \citep{Press89}.

The matrix-based formulation of \eq{LombScargle2} sacrifices some potential computational efficiency in order to attain several advantages:
\begin{enumerate}
  \item It is trivially extended to heteroscedastic and/or correlated measurement noise in the data $y_k$ through appropriate modification of the noise matrix $\Sigma$
  \item It is trivially extended to more sophisticated linear models by appropriately modifying the design matrix $X_\omega$.
  \item It is trivially extended to Tikhonov/L2-regularized likelihoods by adding an appropriate diagonal term to the normal matrix $X_\omega^T\Sigma^{-1}X_\omega$.
\end{enumerate}
In the remainder of this section, we'll explore a few of these modifications and how they affect the periodogram and resulting model fits.

\subsection{Stationary Sinusoid with Floating Mean}
\sectlabel{floating_mean}

@@

As an example of one of these generalizations, consider the {\it generalized Lomb-Scargle} method of \citet{Zechmeister09}. This adjusts the classic normalized periodogram by allowing the mean of the model to be fit alongside the amplitudes:

\begin{equation}
  y(t~|~\omega, \theta) = \theta_0 + \theta_1\sin\omega t + \theta_2\cos\omega t
\end{equation}

This model can be more accurate than the standard periodogram for certain observing cadences and selection functions. While \citet{Zechmeister09} details a formalism to express this floating-mean model (which they call a ``generalized periodogram'') similarly to \eq{LombScargle}, in the matrix formalism all that is required is to add a column of ones to the $X_\omega$ matrix before computing the power via \eq{LombScargle2}.

For well-sampled data, there is little difference between a standard periodogram and a floating-mean periodogram. Where this becomes important is if selection effects or observing cadences cause there to be preferentially more observations at certain phases: an example of this situation is shown in \fig{floating_mean}. The data are drawn from a sinusoid with Gaussian errors, and data with a magnitude fainter than 16 are removed to simulate an observational bias. Because of this bias, the mean of the observed data to be a poor predictor of the true mean, which makes it so that the standard approach fails to find a suitable fit to the data. The floating-mean approach is able to adjust for this, resulting in a periodogram which detects the input period of 0.622 days.

\subsection{Truncated Fourier Models}
\sectlabel{multiterm}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig03.pdf}
  \caption{
    An illustration of a periodogram based on the truncated Fourier model. The data are the same as those in \fig{basic_example}. 
  }
  \figlabel{multiterm_example}
\end{figure}

As mentioned above, the standard periodogram is equivalent to fitting a single-term stationary sinusoidal model to the data. A natural extension is to use a multiple-term sinusoidal model. With $M$ terms, there are $2M + 1$ free parameters, and the model is given by:

\begin{equation}
  y(t|\omega,\theta) = \theta_0 + \sum_{n=1}^M \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right]
\end{equation}

Because this remains a linear model, it can be easily accommodated into the matrix formalism above. For example, an $M = 2$-term model can be constructed by building a design matrix $X_\omega$ with $2M + 1 = 5$ columns:

\begin{equation}
X_\omega^{(2)} = \left[
\begin{array}{ccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & \sin(2\omega t_1) & \cos(2\omega t_1)\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & \sin(2\omega t_2) & \cos(2\omega t_2)\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & \sin(2\omega t_3) & \cos(2\omega t_3)\\
\vdots & \vdots & \vdots & \vdots & \vdots \\
1 & \sin(\omega t_N) & \cos(\omega t_N) & \sin(2\omega t_N) & \cos(2\omega t_N)\\
\end{array}
\right]
\end{equation}
Using $X_\omega^{(2)}$ in \eq{LombScargle2} will result in a two-term periodogram. In general, we can fit an $M$-term periodogram by simply adding terms to the fit. \fig{multiterm_example} shows a few examples of this multiterm Fourier approach as applied to a simulated RR Lyrae light curve constructed from templates derived in \citet{Sesar2010}. There are several important features of this figure that give us insight into the subtleties of this type of fit:

First, we see in the right panel that all three models show a significant spike at the true period of $P_0 = 0.622$. The higher-order models, however, also show a a spike of poer at $P_1 = 2 P_0$: the reason for this is that the fundamental model of the single-term model at $P_0$ is the first harmonic of the 2-term model at $2P_0$, so the higher-order models contain this single-term result.

Second, notice that as the number of terms is increased, the general ``background'' level of the periodogram increases. This is due to the fact that the periodogram is directly related to the $\chi^2$ of the fit. A more complicated model will have the flexibility to conform to the data at {\it all frequencies}, and thus the observed ``power'' is stronger at all frequencies.

\subsection{Regularized Models}
\sectlabel{regularization}
The previous sections raise the question: how complicated a model should we use? If we added many terms to the truncated Fourier series, we might fit the data perfectly, but eventually we would {\it overfit} the data: that is, our model is fitting the noise rather than the signal. This can be avoided by explicitly truncating the series, but another approach is to use a {\it regularization} term to enforce a less complicated model.

A regularization term is an explicit penalty on the magnitude of the model parameters $\theta$, and can take any number of forms. For computational simplicity here we'll use an {\it L2 regularization} (also known as Tikhonov Regularization), which is a quadratic penalty term in the model parameters. This is mathematically equivalent to the Bayesian approach of using a Gaussian prior on the model parameters.

We'll encode our regularization in the matrix $\Lambda = {\rm diag}([\lambda_1, \lambda_2 \cdots \lambda_M])$ for a model with $M$ parameters, and our expression for $\chi^2$ becomes
\begin{equation}
  \eqlabel{chi2reg}
  \chi_\Lambda^2(\omega) = (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta) + \theta^T\Lambda\theta
\end{equation}
Minimizing this $\chi^2$, solving for $\theta$, and plugging into the expression for $P_N$ gives us the regularized counterpart of \eq{LombScargle2}:
\begin{equation}
  \eqlabel{LombScargleReg}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega + \Lambda]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
Note that the effect of this regularization term is to add a diagonal term to the normal matrix $X_\omega^T\Sigma^{-1}X_\omega$, which has the additional feature that it corrects ill-posed models. This feature of the regularization will become important below.

\todo{Show a regularized model?}

\section{A Multiple-Band Model}

To compute a periodogram for multi-band data, we'll take advantage of many of the nice features of the matrix form of the normalized periodogram, which allows us to specify an arbitrary linear model covering our data. We will construct a multi-term Fourier model with the following features:
\begin{enumerate}
  \item An $M_{base}$-term truncated Fourier fit which models a latent parameter, which we'll call the ``overall variability''.
  \item An $M_{band}$-term truncated Fourier fit which models the residual of each band from the overall variability.
\end{enumerate}
The total number of parameters for $K$ filters is then $M_K = (2M_{base} + 1) + K(2M_{band} + 1)$. The model looks as follows, for a band $b$:
\begin{eqnarray}
  y(t|\omega,\theta,b) = &\theta_0 + \sum_{n=1}^{M_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right]& +\\ 
  &\theta^{(b)}_0 + \sum_{n=1}^{M_{band}} \left[\theta^{(b)}_{2n - 1}\sin(n\omega t) + \theta^{(b)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(b)}$ are fit within each band.

We can build the matrix form of this model by building a sparse $X_{\omega}$ matrix with $M$ columns. Each row corresponds to a single observation through a single band; columns corresponding to the base model and the observation band will have nonzero entries; all other columns will be filled with zeros. For example, the $X_\omega$ matrix with $N_{base}=1$ and $N_{band}=0$ (that is, a single sinusoid model with independent offsets to each band) will look as follows:

\begin{equation}
X_\omega^{(1,0)} = \left[
\begin{array}{cccccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & 1 & 0 & 0 & 0 & 0\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & 0 & 1 & 0 & 0 & 0\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & 0 & 0 & 0 & 1 & 0\\
1 & \sin(\omega t_4) & \cos(\omega t_4) & 0 & 0 & 1 & 0 & 0\\
\vdots & \vdots & \vdots & & & \vdots & &\\
1 & \sin(\omega t_N) & \cos(\omega t_N) & 0 & 1 & 0 & 0 & 0\\
\end{array}
\right]
\end{equation}
Here the nonzero entries of the final five columns indicate the band observed at the given time: the first row is a $u$-band measurement, the second is a $g$-band, the third is a $i$-band, etc.

>>>>

\title{Periodograms for Multi-band Astronomical Time Series}

@@

The detection and quantification of periodicity in time-varying signals is an important area of data analysis within modern time-domain astronomical surveys.
For evenly-spaced data, the {\it Schuster periodogram}, introduced in 1905, gives a quantitative measure of the periodicity of data as a function of the angular frequency $\omega$. For data $\{y_k\}_{k=1}^N$ measured at equal intervals $t_k = t_0 + k\Delta t$, the Schuster periodogram, which measures the spectral power as a function of the angular frequency, is given by
\begin{equation}
  \eqlabel{Schuster}
  C(\omega) = \frac{1}{N}\left| \sum_{k=1}^N y_k e^{i\omega t_k} \right|^2,
\end{equation}
and can be computed very efficiently using the Fast Fourier Transform.

Because astronomical observing cadences are rarely so uniform, many have looked at extending the ideas behind the periodogram to work with unevenly-sampled data. Most famously, \citet{Lomb76} and \citet{Scargle82} extended earlier work to define the {\it normalized periodogram}:
\begin{equation}
  \eqlabel{LombScargle}
  P_N(\omega) = \frac{1}{2\sigma^2}\left[
    \frac{\left[\sum_k(y_k - \bar{d})\cos\omega(t_k - \tau)\right]^2}
    {\sum_k \cos^2\omega(t_k - \tau)}
    +
    \frac{\left[\sum_k(y_k - \bar{d})\sin\omega(t_k - \tau)\right]^2}
    {\sum_k \sin^2\omega(t_k - \tau)}
\right],
\end{equation}
where $\bar{d}$ is the mean and $\sigma^2$ is the variance of the data $\{y_k\}$, and $\tau$ is the time-offset which makes $P_N(\omega)$ independent of a translation in $t$ \citep[See][for an in-depth discussion]{NumRec}. \citet{Lomb76} showed that this offset has a deeper effect: namely, it makes $P_N$ identical to the estimate of harmonic content given a least-squares fit to a single-component sinusoidal model,

@@

This long-recognized connection between spectral power and least squares fitting methods was solidified by \citet{Jaynes87}, who demonstrated that the normalized periodogram of Lomb and Scargle is a sufficient statistic for inferences about a stationary-frequency signal in the presence of Gaussian noise. Building on this result, \citet{Bretthorst88} explored the extension of these methods to more complicated models with multiple frequency terms, nonstationary frequencies, and other more sophisticated models within a Bayesian framework.

@@

In this section we will give a quick quantitative introduction to the least squares fitting formulation of the normalized periodogram of \eq{LombScargle}. We'll denote our $N$ observed data points as
\begin{equation}
  D = \{t_k, y_k, \sigma_k\}_{k=1}^N
\end{equation}
where $t_k$ is the time of observation, $y_k$ is the observed value (typically a magnitude), and $\sigma_k$ describes the Gaussian errors on each value. Without loss of generality we will assume that the data $y_k$ are centered such that the measurements within each band satisfy
\begin{equation}
  \eqlabel{ycentered}
  \frac{\sum_k w_ky_k}{\sum_k w_k} = 0
\end{equation}
where the weights are $w_k = \sigma_k^{-2}$.

@@

The normalized periodogram of \eq{LombScargle} can be derived from the normalized $\chi^2$ of the best-fit single-term stationary sinusoidal model given in \eq{SingleModel}. To make the problem linear, we can re-express the model in terms of the parameter vector $\theta = [A\cos\phi, A\sin\phi]$ so that our model is
\begin{equation}
  \eqlabel{simplemodel}
  y(t|\omega,\theta) = \theta_1\sin(\omega t) + \theta_2\cos(\omega t).
\end{equation}
We can find the maximum likelihood estimate of the parameters $\theta$ by minimizing the $\chi^2$ of the model, which is given by
\begin{equation}
  \chi^2(\omega) = \sum_k \frac{[y_k - y(t_k|\omega,\theta)]^2}{2\sigma_k^2}.
\end{equation}
For the single-term Fourier model, it can be shown \citep[See, e.g.][]{ICVG2014} that
\begin{equation}
  \eqlabel{chi2PN}
  \chi_{min}^2(\omega) = \chi^2_0[1 - P_N(\omega)]
\end{equation}
where $P_N(\omega)$ is the normalized periodogram given in \eq{LombScargle} and $\chi^2_0$ is the reference $\chi^2$ for a constant model, which due to the assumption in \eq{ycentered} is simply $\chi^2_0 = \sum_k (y_k/\sigma_k)^2$.

\subsubsection{Matrix Formalism}
The expressions related to the stationary sinusoid model can be expressed more compactly by defining the following matrices:
\begin{equation}
X_\omega = \left[
\begin{array}{cc}
\sin(\omega t_1) & \cos(\omega t_1)\\
\sin(\omega t_2) & \cos(\omega t_2)\\
\vdots & \vdots \\
\sin(\omega t_N) & \cos(\omega t_N)\\
\end{array}
\right];~~
y = \left[
\begin{array}{c}
y_1 \\
y_2\\
\vdots \\
y_N\\
\end{array}
\right];~~
\Sigma = \left[
\begin{array}{cccc}
\sigma_1^2 & 0 &  \cdots & 0\\
0 & \sigma_2^2 &  \cdots & 0\\
\vdots & \vdots &  \ddots & \vdots\\
0 & 0 &  \cdots & \sigma_N^2
\end{array}
\right]
\end{equation}
With these definitions, the model in \eq{simplemodel} can be expressed as a simple linear product: $y(t|\omega,\theta) = X_\omega\theta$, and the model and reference $\chi^2$ can be written

\begin{eqnarray}
  \chi^2(\omega) &=& (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta)\\
  \chi^2_0 &=& y^T \Sigma^{-1} y
\end{eqnarray}
The expression for the normalized periodogram can be computed by finding the value of $\theta$ which minizes $\chi^2(\omega)$ via standard methods and plugging the result into \eq{chi2PN} to find
\begin{equation}
  \eqlabel{LombScargle2}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
Take note that this expression is equivalent to \eq{LombScargle} in the homoscedastic case with $\Sigma \propto \sigma^2 I$.

@@

As an example of the standard periodogram in action, we perform a simple single-band harmonic analysis of simulated $r$-band observations of an RR Lyrae light curve, based on empirical templates derived in \citet{Sesar2010} (\fig{basic_example}). The observations are of a star with a period of 0.622 days, and take place on 60 random nights over a 6-month period, as seen in the left panel.

The upper-right panel shows the normalized periodogram for this source as a function of period. While the power does peak at the true period of 0.622 days, an aliasing effect is readily apparent near $P=0.38$. This additional peak is due to beat frequency between the true period $P$ and the observing cadence of $\sim 1$ day. This beat frequency is the first in a large sequence: for nightly observations, we'd expect to find excess power at periods That is, for nightly observations, we'd expect excess power to be found at frequencies $P_n = P / (1 + nP)$ days, for any positive or negative integer $n$. The strong alias in \fig{basic_example} corresponds to the $n=1$ beat period $P_n=0.383$. Though it is possible to carefully correct for such aliasing via the estimated window function \citep[e.g.][]{Roberts87}, we'll ignore this detail in the current work.

The lower-right panel of \fig{basic_example} shows the maximum likelihood interpretation of this periodogram: it is a measure of the normalized $\chi^2$ for a single-term sinusoidal model. Here we visualize the data from the left panel, but folded as a function of phase with the best-fit single-term model over-plot. Here its apparent that the single-term model is biased: RR Lyrae light curves are, in general, much more complicated than a simple sinusoid. Nevertheless, the simplistic sinusoidal model does well is able to recover the correct frequency to a high degree of accuracy (roughly related to the width of the peak) and significance (roughly related to the height of the peak). For a more complete introduction to and discussion of the single-term normalized periodogram, refer to, e.g. \citet{Bretthorst88} or \citet{ICVG2014}.

\subsection{Extending the Periodogram}
We have shown two forms of the classic normalized periodogram: \eq{LombScargle} and \eq{LombScargle2}. Though the two expressions are equivalent, they each have distinct advantages. The expression in \eq{LombScargle} avoids the explicit construction of a matrix, and thus can be computed very efficiently. Furthermore, through some computational tricks taking advantage of the Fast Fourier Transform, expressions of the form of \eq{LombScargle} can be evaluated exactly for $N$ frequencies in $\mathcal{O}[\log{N}]$ time \citep{Press89}.

The matrix-based formulation of \eq{LombScargle2}, though slower than the Fourier-derived formulation, is a more general expression and allows several advantages:
\begin{enumerate}
  \item It is trivially extended to heteroscedastic and/or correlated measurement noise in the data $y_k$ through appropriate modification of the noise covariance matrix $\Sigma$
  \item It is trivially extended to more sophisticated linear models by appropriately modifying the design matrix $X_\omega$.
  \item It is trivially extended to include Tikhonov/L2-regularization terms by adding an appropriate diagonal term to the normal matrix $X_\omega^T\Sigma^{-1}X_\omega$.
\end{enumerate}
In the remainder of this section, we'll explore a few of these modifications and how they affect the periodogram and resulting model fits.

\subsubsection{Stationary Sinusoid with Floating Mean}
\sectlabel{floating_mean}

@@

As an example of one of these generalizations, consider the {\it generalized Lomb-Scargle} method of \citet{Zechmeister09}. This adjusts the classic normalized periodogram by allowing the mean of the model to be fit alongside the amplitudes:
\begin{equation}
  y(t~|~\omega, \theta) = \theta_0 + \theta_1\sin\omega t + \theta_2\cos\omega t
\end{equation}
This model can be more accurate than the standard periodogram for certain observing cadences and selection functions. \citet{Zechmeister09} detail the modifications required to the harmonic formalism of \eq{LombScargle} to allow the mean to float in the model. In the matrix formalism, the modification is much more straightforward: all that is required is to add a column of ones to the $X_\omega$ matrix before computing the power via \eq{LombScargle2}.

For well-sampled data, there is usually very little difference between a standard periodogram and a floating-mean periodogram. Where this becomes important is if selection effects or observing cadences cause there to be preferentially more observations at certain phases of the light curve: a toy example demonstrating this situation is shown in \fig{floating_mean}. The data are drawn from a sinusoid with Gaussian errors, and data with a magnitude fainter than 16 are removed to simulate an observational bias (right panel). Because of this selection effect, the mean of the observed data are a poor predictor of the true mean, causing the standard fixed-mean method to poorly fit the data. The floating-mean approach is able to automatically adjust for this bias, resulting in a periodogram which readily detects the input period of 0.622 days (lower-right panel).

\subsubsection{Truncated Fourier Models}
\sectlabel{multiterm}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig03.pdf}
  \caption{
    An illustration of a periodogram based on the truncated Fourier model.
    The data are the same as those in \fig{basic_example}. Note that the
    higher-order models will show periodogram peaks at multiples of the true
    fundamental frequency $P_0$: this is because for integer $n$ less than
    the number of Fourier terms in the model, $P_0$ is a higher harmonic of
    the model at $P=nP_0$.
  }
  \figlabel{multiterm_example}
\end{figure}

As mentioned above, the standard periodogram is equivalent to fitting a single-term stationary sinusoidal model to the data. A natural extension is to instead use a multiple-term sinusoidal model, with frequencies at integer multiples of the fundamental frequency. With $M$ terms, there are $2M + 1$ free parameters, and the model is given by
\begin{equation}
  y(t|\omega,\theta) = \theta_0 + \sum_{n=1}^M \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right].
\end{equation}
Because this remains a linear model, it can be easily accommodated into the matrix formalism above. For example, an $M = 2$-term floating-mean model can be constructed by building a design matrix $X_\omega$ with $2M + 1 = 5$ columns:
\begin{equation}
X_\omega^{(2)} = \left[
\begin{array}{ccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & \sin(2\omega t_1) & \cos(2\omega t_1)\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & \sin(2\omega t_2) & \cos(2\omega t_2)\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & \sin(2\omega t_3) & \cos(2\omega t_3)\\
\vdots & \vdots & \vdots & \vdots & \vdots \\
1 & \sin(\omega t_N) & \cos(\omega t_N) & \sin(2\omega t_N) & \cos(2\omega t_N)\\
\end{array}
\right]
\end{equation}
Computing the power via \eq{LombScargle2} using $X_\omega^{(2)}$ will give the two-term periodogram. As $M$ grows, the size of the design matrix grows, but the periodogram can be computed in the same manner. \fig{multiterm_example} shows a few examples of this multiterm Fourier approach as applied to the simulated RR Lyrae light curve from \fig{basic_example}. There are several important features of this figure that give us insight into the subtleties of this type of multiterm fit.

First, we see in the right panel that all three models show a significant spike at the true period of $P_0 = 0.622$. The higher-order models, however, also show a a spike in power at $P_1 = 2 P_0$: the reason for this is that for and $M>1$-term model, the period $P_0$ is the first harmonic of a model with fundamental frequency $2P_0$, and the higher-order models contain the single-period result.

Second, notice that as the number of terms is increased, the general ``background'' level of the periodogram increases. This is due to the fact that the periodogram is directly related to the $\chi^2$ of the fit at each frequency. Because of the increased model complexity of a higher-order model, it is able to better fit the data at all periods, not just the true period. Thus the observed ``power'' of a higher-order model will be everywhere higher than the power of a lower-order model.

\subsubsection{Regularized Models}
\sectlabel{regularization}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig04.pdf}
  \caption{
    The effect of regularization on a high-order model. The data is the same as
    those in \fig{basic_example}. We fit a 20-term truncated Fourier model to
    the data, with and without a regularization term. Without regularization,
    the model oscillates widely to fit the noise in the data. The
    regularization term effectively damps the higher-order Fourier modes and
    removes this oscillating behavior, leading to a model which is much more
    robust to noise in the data.
  }
  \figlabel{regularized_example}
\end{figure}

The previous sections raise the question: how complicated a model should we use? It is easily shown that as we add more terms to the fit, the model will come closer and closer to the observed data. When the number of model parameters equals the number of data points, we can expect to fit the data {\it perfectly} (though in most cases numerical inaccuracies in the fit prevent a truly perfect fit). For very high-order models, the result become very sensitive to the noise in the data, and we say we have {\it over-fit} the data. This can be addressed by explicitly truncating the series, but we can also use a {\it regularization} term to mathematically ensure a less complicated model.

A regularization term is an explicit penalty on the magnitude of the model parameters $\theta$, and can take a number of forms. For computational simplicity here we'll use an {\it L2 regularization} (also known as Tikhonov Regularization), which is a quadratic penalty term in the model parameters added to the $\chi^2$. Mathematically, this is equivalent to the Bayesian approach of using a Gaussian prior on the model parameters.

We'll encode our regularization in the matrix $\Lambda = {\rm diag}([\lambda_1, \lambda_2 \cdots \lambda_M])$ for a model with $M$ parameters, and our expression for $\chi^2$ becomes
\begin{equation}
  \eqlabel{chi2reg}
  \chi_\Lambda^2(\omega) = (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta) + \theta^T\Lambda\theta
\end{equation}
Minimizing this $\chi^2$, solving for $\theta$, and plugging into the expression for $P_N$ gives us the regularized counterpart of \eq{LombScargle2}:
\begin{equation}
  \eqlabel{LombScargleReg}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega + \Lambda]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
Notice that the effect of this regularization term is to add a diagonal penalty term to the normal matrix $X_\omega^T\Sigma^{-1}X_\omega$, which has the additional feature that it can correct ill-posed models where the normal matrix is not invertible. This feature of the regularization will become important below.

In \fig{regularized_example}, we compare a regularized and unregularized 20-term truncated Fourier model on our simulated RR Lyrae light curve. We use $\lambda = 0$ on the offset term, and make $\lambda_j$ progressively larger for each harmonic component. The result of the regularized model is less over-fitting to the input data, and a more distinct peaks in the associated periodogram.

\section{A Multiple-Band Model}
To compute a periodogram for multi-band data, we'll take advantage of many of the nice features of the matrix form of the normalized periodogram, which allows us to specify an arbitrary linear model covering our data. We will construct a multi-term Fourier model with the following features:
\begin{enumerate}
  \item An $M_{base}$-term truncated Fourier fit which models a latent parameter, which we'll call the ``overall variability''.
  \item An $M_{band}$-term truncated Fourier fit which models the residual of each band from this overall variability.
\end{enumerate}
The total number of parameters for $K$ filters is then $M_K = (2M_{base} + 1) + K(2M_{band} + 1)$. As a result, for each band $b$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  y_b(t|\omega,\theta) = &\theta_0 + \sum_{n=1}^{M_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right]& +\\ 
  &\theta^{(b)}_0 + \sum_{n=1}^{M_{band}} \left[\theta^{(b)}_{2n - 1}\sin(n\omega t) + \theta^{(b)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(b)}$ are determined individually.
